Fixing Vulnerabilities Hinders Maintainability

### 2020-01-31 paper/icpc20.tex
new version

%%%%% NEEDS REVIEW
As an example, consider the patch of the state-machine implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting TLS messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-$2014$-$3506$\footnote{CVE-$2014$-$3506$
details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506}
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
\emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
\url{https://github.com/openssl/openssl/commit/1250f12613b61758675848f6600ebd914ccd7636}
(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between the two parties. DTLS provides a mechanism for fragmenting
a handshake message over a number of packets, each of which can be transmitted
separately over the internet from one party to another. The problem in
CVE-$2014$-$3506$ is that the size of the handshake message collected by the
receiver is not verified against the maximum size allowed by DTLS. It is then
allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~36). Thus, if an
attacker has access to one of these packages, he can change the
handshake message size and trigger unnecessary memory allocations of large
length which can lead to memory exhaustion that could end in a DoS attack.

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) {
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}
%
%\medskip
%\setcounter{lstannotation}{0}
%\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
%\\CVE-2016-6307 vulnerability for file ssl/statem/statem_lib.c },label={lst:vuln}]
% int tls_get_message_header(SSL *s, int *mt){
% // [snip]
%+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
%+        && !BUF_MEM_grow_clean(s->init_buf, 
%+                                (int)s->s3->tmp.message_size
%+                                + SSL3_HM_HEADER_LENGTH)) {
%+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
%+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
%+          return SUB_STATE_ERROR;
% // [snip]
% }
%\end{lstlisting}

@@

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to return the maximum number of bytes allowed in a DTLS
handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if
the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires it. More
specifically, it checks if the maximum certificate list size requires more length 
than the maximum length (\texttt{max\_len}). If that condition holds, the function 
returns the size of the certificate list, otherwise, it returns the maximum length. 
This function replaces lines $17$--$21$.

~\ref{lst:func5} and~\ref{lst:func4} are guards added to different functions to
ascertain that the message and the fragment length are never higher than the
maximum allowed by DTLS. In particular, \ref{lst:func5} is responsible for patching
the vulnerability.

@@

Although checking the maximum size of a handshake message seems an elementary

@@

maintainability. Even though the developer tries to simplify the
code base by introducing a smaller unit
\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change disrupts

@@

building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
point to the \sloppy\texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline which resulted in a negative impact on the

>>>>

%%%%% NEEDS REVIEW
As an example, consider the patch of the TLS state-machine protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting TLS messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$ \footnote{CVE-$2016$-$6307$
details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
\emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
\url{https://github.com/openssl/openssl/commit/c1ef7c971d0bbf117c3c80f65b5875e2e7b024b1}
(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between the two parties. The problem in
CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
memory for a message is allocated too early, prior to the excessive
message length check which can lead to a Denial of Service through memory
exhaustion if an attacker triggers up a certain message length to be allocated to service a
connection. In this case, there is only a security impact if the application does not call  \texttt{SSL\_free()} in a timely manner in the
event that the connection fails;
or, the application is working in a constrained environment where there
is very little free memory;
or, the attacker initiates multiple connection attempts such that there
are multiple connections in a state where memory has been allocated for
the connection, \texttt{SSL\_free()} has not yet been called, and there is
insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
be transitory because as soon as the connection fails the memory is
subsequently freed again in the \texttt{SSL\_free()} call. However there is an
increased risk during this period of application crashes due to the lack
of memory - which would then mean a more serious Denial of Service.

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
 int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
 // [snip]
  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
        SSL3_HM_HEADER_LENGTH;
- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-      goto err;
   }
 // [snip]

- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func5}!*/  
-     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-     goto err;
- }

\end{lstlisting}
%
%\medskip
%\setcounter{lstannotation}{0}
%\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
%\\CVE-2016-6307 vulnerability for file ssl/statem/statem_lib.c },label={lst:vuln}]
% int tls_get_message_header(SSL *s, int *mt){
% // [snip]
%+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
%+        && !BUF_MEM_grow_clean(s->init_buf, 
%+                                (int)s->s3->tmp.message_size
%+                                + SSL3_HM_HEADER_LENGTH)) {
%+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
%+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
%+          return SUB_STATE_ERROR;
% // [snip]
% }
%\end{lstlisting}

@@

~\ref{lst:func1} \texttt{read\_state\_machine} function abstracts a sub-state-machine
to handle the messages exchanged between peers. When messages are larger than
what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
of this resource was being verified very early in the application's flow.  Listing ~\ref{lst:vuln2} presents those verifications (lines $5$-$7$ and line $11$-$13$).
These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}.

@@

Although checking the maximum size of a message seems an elementary

@@

maintainability. The developer tries to patch the
code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
and disrupts

@@

building maintainable software~\cite{Visser:2016:OREILLY} $-$
the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
